---
title: "Time Series Forecasting with Robyn"
author: "Emiliia Lekhman"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

This report presents a time series forecasting model using the Robyn library in R.

## Load Libraries

```{r, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
library(tidyr)
library(gridExtra)
library(tidyverse)
library(zoo)
library(siMMMulator)
library(forecast)
library(Robyn)
library(prophet)

create_files <- TRUE
```

# Generate Data with siMMMulator

## Step 0 : Define Basic Parameters
```{r, message=FALSE, warning=FALSE}
my_variables <- step_0_define_basic_parameters(years = 2,
                                               channels_impressions = c("Facebook", "TV", "Long_Form_Video"),
                                               channels_clicks = c("Search"),
                                               frequency_of_campaigns = 1,
                                               true_cvr = c(0.001, 0.002, 0.003, 0.002),
                                               revenue_per_conv = 1, 
                                               start_date = "2017/1/1"
)
```

## Step 1 : Simulate Daily Baseline Sales
```{r, message=FALSE, warning=FALSE}
df_baseline <- step_1_create_baseline(
                        my_variables = my_variables,
                        base_p = 500000,
                        trend_p = 1.8,
                        temp_var = 8,
                        temp_coef_mean = 50000,
                        temp_coef_sd = 5000,
                        error_std = 100000)
optional_step_1.5_plot_baseline_sales(df_baseline = df_baseline)
```

## Step 2 : Generate Ad Spend

```{r, message=FALSE, warning=FALSE}
df_ads_step2 <- step_2_ads_spend(
                    my_variables = my_variables,
                    campaign_spend_mean = 329000,
                    campaign_spend_std = 100000,
                    max_min_proportion_on_each_channel <- c(0.15, 0.20,
                                                            0.10, 0.15,
                                                            0.05, 0.20)
                )
optional_step_2.5_plot_ad_spend(df_ads_step2 = df_ads_step2)
```

## Step 3 : Generate Media Variables

```{r, message=FALSE, warning=FALSE}
df_ads_step3 <- step_3_generate_media(
  my_variables = my_variables,
  df_ads_step2 = df_ads_step2,
  true_cpm = c(2, 20, 10, NA),
  true_cpc = c(NA, NA, NA, 0.25),
  mean_noisy_cpm_cpc = c(1, 0.05, 0.01, 0.1),
  std_noisy_cpm_cpc = c(0.01, 0.15, 0.01, 0.01)
)
```

## Step 4 : Generate Noisy CVRs

```{r, message=FALSE, warning=FALSE}
df_ads_step4 <- step_4_generate_cvr(
  my_variables = my_variables,
  df_ads_step3 = df_ads_step3,
  mean_noisy_cvr = c(0, 0.0001, 0.0002, 0), 
  std_noisy_cvr = c(0.001, 0.002, 0.003, 0.002)
  )
```

## Step 5 : Transforming Media Variables

### Step 5a : Pivoting the table to an MMM format

```{r, message=FALSE, warning=FALSE}
df_ads_step5a_before_mmm <- step_5a_pivot_to_mmm_format(
                    my_variables = my_variables,
                    df_ads_step4 = df_ads_step4
                    )
```

### Step 5b : Apply Adstock

```{r, message=FALSE, warning=FALSE}
df_ads_step5b <- step_5b_decay(
  my_variables = my_variables,
  df_ads_step5a_before_mmm = df_ads_step5a_before_mmm,
  true_lambda_decay = c(0.1, 0.2, 0.3, 0.4)
)
```

### Step 5c: Applying Diminishing Returns to Media Variables

```{r, message=FALSE, warning=FALSE}
df_ads_step5c <- step_5c_diminishing_returns(
  my_variables = my_variables,
  df_ads_step5b = df_ads_step5b,
  alpha_saturation = c(2, 2, 2, 2),
  gamma_saturation = c(0.1, 0.2, 0.3, 0.2)
)
```

## Step 6 : Calculating Conversions

```{r, message=FALSE, warning=FALSE}
df_ads_step6 <- step_6_calculating_conversions(
  my_variables = my_variables,
  df_ads_step5c = df_ads_step5c
)
```

## Step 7 : Expanded Data Frame

```{r, message=FALSE, warning=FALSE}
df_ads_step7 <- step_7_expanded_df(
  my_variables = my_variables,
  df_ads_step6 = df_ads_step6,
  df_baseline = df_baseline
)
```

## Step 8 : Calculating ROI

```{r, message=FALSE, warning=FALSE}
step_8_calculate_roi(
  my_variables = my_variables,
  df_ads_step7 = df_ads_step7
  )
```

## Step 9 : Get Final Data Frame

```{r, message=FALSE, warning=FALSE}
list_of_df_final <- step_9_final_df(
  my_variables = my_variables,
  df_ads_step7 = df_ads_step7
)
daily_df <- list_of_df_final[[1]]
optional_step_9.5_plot_final_df(df_final = list_of_df_final[[1]]) # for daily data
head(daily_df)
tail(daily_df)
names(daily_df)
```

# Time Series Forecasting

### Step 1 : Input

```{r, message=FALSE, warning=FALSE}
robyn_directory <- "./robyn"
```

```{r, message=FALSE, warning=FALSE}
InputCollect <- robyn_inputs(
  dt_input = daily_df,
  dt_holidays = dt_prophet_holidays,
  date_var = "DATE", # date format must be "2020-01-01"
  dep_var = "total_revenue", # there should be only one dependent variable
  dep_var_type = "revenue", # "revenue" (ROI) or "conversion" (CPA)
  prophet_vars = c("trend", "season", "weekday", "monthly", "holiday"), # "trend","season", "weekday" & "holiday"
  prophet_country = "DE", # input country code. Check: dt_prophet_holidays
  # context_vars = c("competitor_sales_B", "events"), # e.g. competitors, discount, unemployment etc
  paid_media_spends = c("spend_Facebook", "spend_TV", "spend_Long_Form_Video", "spend_Search"), # mandatory input
  paid_media_vars = c("impressions_Facebook", "impressions_TV", "impressions_Long_Form_Video", "clicks_Search"), # mandatory.
  # paid_media_vars must have same order as paid_media_spends. Use media exposure metrics like
  # impressions, GRP etc. If not applicable, use spend instead.
  # organic_vars = "newsletter", # marketing activity without media spend
  # factor_vars = c("events"), # force variables in context_vars or organic_vars to be categorical
  window_start = "2017-01-01",
  window_end = "2018-12-31",
  adstock = "geometric" # geometric, weibull_cdf or weibull_pdf.
)
print(InputCollect)
```

```{r, message=FALSE, warning=FALSE}
hyper_names(adstock = InputCollect$adstock, all_media = InputCollect$all_media)
```

```{r, message=FALSE, warning=FALSE}
plot_adstock(plot = TRUE)
plot_saturation(plot = TRUE)
```

```{r, message=FALSE, warning=FALSE}
hyper_limits()

# Example hyperparameters ranges for Geometric adstock
hyperparameters <- list(
  spend_Facebook_alphas = c(0.5, 3),
  spend_Facebook_gammas = c(0.3, 1),
  spend_Facebook_thetas = c(0, 0.3),
  spend_TV_alphas = c(0.5, 3),
  spend_TV_gammas = c(0.3, 1),
  spend_TV_thetas = c(0.1, 0.4),
  spend_Long_Form_Video_alphas = c(0.5, 3),
  spend_Long_Form_Video_gammas = c(0.3, 1),
  spend_Long_Form_Video_thetas = c(0.3, 0.8),
  spend_Search_alphas = c(0.5, 3),
  spend_Search_gammas = c(0.3, 1),
  spend_Search_thetas = c(0, 0.3),
  train_size = c(0.5, 0.8)
)
```

```{r, message=FALSE, warning=FALSE}
InputCollect <- robyn_inputs(InputCollect = InputCollect, hyperparameters = hyperparameters)
print(InputCollect)
```

### Step 2 : Build initial model

```{r, message=FALSE, warning=FALSE}
Sys.setenv(KMP_DUPLICATE_LIB_OK = "TRUE")

OutputModels <- robyn_run(
  InputCollect = InputCollect, # feed in all model specification
  cores = NULL, # NULL defaults to (max available - 1)
  iterations = 10, # 2000 recommended for the dummy dataset with no calibration
  trials = 5, # 5 recommended for the dummy dataset
  ts_validation = TRUE, # 3-way-split time series for NRMSE validation.
  add_penalty_factor = FALSE # Experimental feature. Use with caution.
)
```

```{r, message=FALSE, warning=FALSE}
print(OutputModels)

OutputModels$convergence$moo_distrb_plot
OutputModels$convergence$moo_cloud_plot

if (OutputModels$ts_validation) OutputModels$ts_validation_plot
```

```{r, message=FALSE, warning=FALSE}
OutputCollect <- robyn_outputs(
  InputCollect, OutputModels,
  pareto_fronts = "auto", # automatically pick how many pareto-fronts to fill min_candidates (100)
  min_candidates = 2, # top pareto models for clustering. Default to 100
  # calibration_constraint = 0.1, # range c(0.01, 0.1) & default at 0.1
  csv_out = "pareto", # "pareto", "all", or NULL (for none)
  clusters = FALSE, # Set to TRUE to cluster similar models by ROAS. See ?robyn_clusters
  export = create_files, # this will create files locally
  plot_folder = robyn_directory, # path for plots exports and files creation
  plot_pareto = create_files # Set to FALSE to deactivate plotting and saving model one-pagers
)
print(OutputCollect)
```

```{r, message=FALSE, warning=FALSE}
select_model <- "3_2_5" # Pick one of the models from OutputCollect to proceed

#### Version >=3.7.1: JSON export and import (faster and lighter than RDS files)
ExportedModel <- robyn_write(InputCollect, OutputCollect, select_model, export = create_files)
print(ExportedModel)

# To plot any model's one-pager:
myOnePager <- robyn_onepagers(InputCollect, OutputCollect, select_model, export = FALSE)
```

```{r, message=FALSE, warning=FALSE}
InputCollect$paid_media_spends
