---
title: "Time Series Forecasting with Prophet"
author: "Emiliia Lekhman"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

This report presents a time series forecasting model using the Prophet library in R.

## Load Libraries

```{r, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
library(tidyr)
library(gridExtra)
library(tidyverse)
library(zoo)
library(forecast)
library(prophet)
library(Metrics)
```

# Load data
```{r, message=FALSE, warning=FALSE}
# Load datasets
train <- read_csv("data/train.csv")
test <- read_csv("data/test.csv")
stores <- read_csv("data/stores.csv")
oil <- read_csv("data/oil.csv")
holidays <- read_csv("data/holidays_events.csv")
```

```{r}
length(unique(rownames(train)))
```

```{r, message=FALSE, warning=FALSE}
head(train)
```

```{r, message=FALSE, warning=FALSE}
str(train)
str(test)
str(stores)
str(oil)
str(holidays)
```

# Data Preprocessing

```{r, message=FALSE, warning=FALSE}
# Convert date to date format
train$date <- as.Date(train$date)
test$date <- as.Date(test$date)
oil$date <- as.Date(oil$date)
holidays$date <- as.Date(holidays$date)
```

```{r}
length(unique(rownames(train)))
```

```{r, message=FALSE, warning=FALSE}
# Merge train datasets
train <- train %>% left_join(stores, by = "store_nbr")
train <- train %>% left_join(oil, by = "date")
train <- train %>% left_join(holidays, by = "date")
names(train)[names(train) == "type.x"] <- "type.store"
names(train)[names(train) == "type.y"] <- "type.holiday"
head(train)
```

```{r, message=FALSE, warning=FALSE}
# Merge test datasets
test <- test %>% left_join(stores, by = "store_nbr")
test <- test %>% left_join(oil, by = "date")
test <- test %>% left_join(holidays, by = "date")
names(test)[names(test) == "type.x"] <- "type.store"
names(test)[names(test) == "type.y"] <- "type.holiday"
head(test)
```

```{r, message=FALSE, warning=FALSE}
duplicates <- train %>%
  group_by(date, store_nbr, family) %>%
  summarise(count = n()) %>%
  filter(count > 1)

# Print the result
if (nrow(duplicates) == 0) {
  print("No duplicate combinations of store_nbr and family for each date.")
} else {
  print("There are duplicate combinations of store_nbr and family for some dates.")
  print(duplicates)
}
```

```{r, message=FALSE, warning=FALSE}
store_city_check <- train %>%
  group_by(store_nbr) %>%
  summarise(unique_cities = n_distinct(city)) %>%
  filter(unique_cities > 1)

# Print the result
if (nrow(store_city_check) == 0) {
  print("Each store_nbr is associated with only one unique city.")
} else {
  print("Some store_nbr values are associated with multiple cities.")
  print(store_city_check)
}
```
 
```{r, message=FALSE, warning=FALSE}
# Check missing values in train dataset
sapply(train, function(x) sum(is.na(x)))
```

```{r, message=FALSE, warning=FALSE}
# Check missing values in test dataset
sapply(test, function(x) sum(is.na(x)))
```

```{r, message=FALSE, warning=FALSE}
# Fill missing values
train$onpromotion[is.na(train$onpromotion)] <- 0
train$dcoilwtico <- na.fill(train$dcoilwtico, fill = "extend")
train$dcoilwtico <- na.approx(train$dcoilwtico)
train$transferred[is.na(train$transferred)] <- "False"
train$locale_name[is.na(train$locale_name)] <- "Unknown"
train$locale[is.na(train$locale)] <- "Unknown"
train$type.holiday[is.na(train$type.holiday)] <- "Unknown"
train$description[is.na(train$description)] <- "Unknown"
```

```{r, message=FALSE, warning=FALSE}
# Fill missing values in test dataset
test$onpromotion[is.na(test$onpromotion)] <- 0
test$dcoilwtico <- na.fill(test$dcoilwtico, fill = "extend")
test$dcoilwtico <- na.approx(test$dcoilwtico)
test$transferred[is.na(test$transferred)] <- "False"
test$locale_name[is.na(test$locale_name)] <- "Unknown"
test$locale[is.na(test$locale)] <- "Unknown"
test$type.holiday[is.na(test$type.holiday)] <- "Unknown"
test$description[is.na(test$description)] <- "Unknown"
```


```{r, message=FALSE, warning=FALSE}
# Check missing values again
sapply(train, function(x) sum(is.na(x)))
sapply(test, function(x) sum(is.na(x)))
```


# Exploratory Data Analysis

## General Sales Trends

```{r, message=FALSE, warning=FALSE}
train %>%
  group_by(date) %>%
  summarize(total_sales = sum(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = date, y = total_sales)) +
  geom_line(color = "blue") +
  labs(title = "Total Sales Over Time", x = "Date", y = "Total Sales")
```

## Sales by Store Family

```{r, message=FALSE, warning=FALSE}
train %>%
  group_by(family) %>%
  summarize(total_sales = sum(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = family, y = total_sales)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Total Sales by Family", x = "Family", y = "Total Sales") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate text
```

## Sales by Holiday Type

```{r, message=FALSE, warning=FALSE}
train %>%
  filter(type.holiday != "Unknown") %>%  # Exclude "Unknown" holiday type
  group_by(type.holiday) %>%
  summarize(total_sales = sum(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = type.holiday, y = total_sales)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Total Sales by Holiday Type", x = "Holiday Type", y = "Total Sales")
```

## Sales by Locale

```{r, message=FALSE, warning=FALSE}
train %>%
  filter(locale_name != "Unknown") %>%
  group_by(locale_name) %>%
  summarize(total_sales = sum(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = locale_name, y = total_sales)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Total Sales by Locale", x = "Locale", y = "Total Sales") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Oil Price Trends

```{r, message=FALSE, warning=FALSE}
train %>%
  ggplot(aes(x = date, y = dcoilwtico)) +
  geom_line(color = "blue") +
  labs(title = "Oil Price Over Time", x = "Date", y = "Oil Price")
```

## Sales vs. Promotions

```{r, message=FALSE, warning=FALSE}
train %>%
  group_by(onpromotion) %>%
  summarize(avg_sales = mean(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = onpromotion, y = avg_sales)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Average Sales by Number of Promotions", x = "Number of Promotions", y = "Average Sales")
```

## Seasonality Analysis

```{r, message=FALSE, warning=FALSE}
train %>%
  mutate(dayOfWeek = wday(date, label = TRUE)) %>%
  group_by(dayOfWeek) %>%
  summarize(avg_sales = mean(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = dayOfWeek, y = avg_sales)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Average Sales by Day of Week", x = "Day of the Week", y = "Average Sales")
```

## Impact of Holidays on Sales

```{r, message=FALSE, warning=FALSE}
train %>%
  filter(type.holiday == "Holiday" | type.holiday == "Event") %>%
  group_by(date, type.holiday) %>%
  summarize(total_sales = sum(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = date, y = total_sales, color = type.holiday)) +
  geom_line() +
  labs(title = "Total Sales by Holiday Type", x = "Date", y = "Total Sales")
```

# Time Series Analysis

## Time Series Decomposition

```{r, message=FALSE, warning=FALSE}
  # Aggregate sales by date
  daily_sales <- train %>%
    group_by(date) %>%
    summarize(total_sales = sum(sales))

  # Decompose the time series
  decomposition <- stl(ts(daily_sales$total_sales, frequency = 7), s.window = "periodic")
  plot(decomposition)
```

## ACF and PACF Plots

```{r, message=FALSE, warning=FALSE}
  # ACF and PACF plots
  acf(daily_sales$total_sales, main = "ACF Plot")
  pacf(daily_sales$total_sales, main = "PACF Plot")
```

# Time Series Forecasting  

## Prepare Data for Prophet

```{r, message=FALSE, warning=FALSE}
train$type.holiday <- as.factor(train$type.holiday)
train$store_nbr <- as.factor(train$store_nbr)
train$family <- as.factor(train$family)
train$city <- as.factor(train$city)
train$state <- as.factor(train$state)
train$type.store <- as.factor(train$type.store)
train$cluster <- as.factor(train$cluster)
train$type.holiday <- as.factor(train$type.holiday)
train$locale <- as.factor(train$locale)
train$locale_name <- as.factor(train$locale_name)
train$transferred <- as.factor(train$transferred)
train$description <- as.factor(train$description)

test$type.holiday <- as.factor(test$type.holiday)
test$store_nbr <- as.factor(test$store_nbr)
test$family <- as.factor(test$family)
test$city <- as.factor(test$city)
test$state <- as.factor(test$state)
test$type.store <- as.factor(test$type.store)
test$cluster <- as.factor(test$cluster)
test$type.holiday <- as.factor(test$type.holiday)
test$locale <- as.factor(test$locale)
test$locale_name <- as.factor(test$locale_name)
test$transferred <- as.factor(test$transferred)
test$description <- as.factor(test$description)

train_prophet <- train %>%
  rename(ds = date, y = sales)

str(train_prophet)
head(train_prophet)
tail(train_prophet)
```

## Train and Validation Split

```{r, message=FALSE, warning=FALSE}
split_ratio <- 0.9
train_size <- floor(split_ratio * nrow(train_prophet))

train_data <- train_prophet[1:train_size, ]
val_data <- train_prophet[(train_size + 1):nrow(train_prophet), ]

dim(train_data)
dim(val_data)
dim(test)
```


## Fit the baseline model

```{r, message=FALSE, warning=FALSE}
# model_baseline <- prophet(train_data)
# saveRDS(model_baseline, file="model_baseline.RDS")
```

## Fit the all_features model

```{r, message=FALSE, warning=FALSE}
# model_all_features <- prophet()
# model_all_features <- add_regressor(model_all_features, 'store_nbr', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'family', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'onpromotion')
# model_all_features <- add_regressor(model_all_features, 'city', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'state', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'type.store', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'cluster', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'dcoilwtico')
# model_all_features <- add_regressor(model_all_features, 'type.holiday', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'locale', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'locale_name', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'description', standardize = FALSE)
# model_all_features <- add_regressor(model_all_features, 'transferred', standardize = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# model_all_features <- fit.prophet(model_all_features, train_data)
# saveRDS(model_all_features, file="model_all_features.RDS")
```


## Fit the some_features model

```{r, message=FALSE, warning=FALSE}
# model_some_features <- prophet()
# model_some_features <- add_regressor(model_some_features, 'family', standardize = FALSE)
# model_some_features <- add_regressor(model_some_features, 'onpromotion')
# model_some_features <- add_regressor(model_some_features, 'city', standardize = FALSE)
# model_some_features <- add_regressor(model_some_features, 'type.store', standardize = FALSE)
# model_some_features <- add_regressor(model_some_features, 'dcoilwtico')
# model_some_features <- add_regressor(model_some_features, 'type.holiday', standardize = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# model_some_features <- fit.prophet(model_some_features, train_data)
# saveRDS(model_some_features, file="model_some_features.RDS")
```

## Fit the models with parameters

```{r, message=FALSE, warning=FALSE}
# # Prepare holidays data (assuming holidays data is already included in `train_data`)
# holidays <- train_data %>%
#   filter(type.holiday != "Unknown") %>%  # Filter out non-holiday rows
#   select(ds = ds, holiday = type.holiday) %>%  # Rename columns to match Prophet's expected format
#   distinct()

# # Define a changepoint for the earthquake date
# earthquake_changepoint <- as.Date("2016-04-16")

# # Set up the Prophet model with holidays, additional regressors, and a custom changepoint
# model_with_params <- prophet(
#   holidays = holidays, 
#   yearly.seasonality = TRUE, 
#   weekly.seasonality = TRUE, 
#   daily.seasonality = FALSE, 
#   seasonality.mode = 'additive', 
#   changepoints = c(earthquake_changepoint)  # Adding earthquake as a custom changepoint
# )

# # Add each additional feature in the dataset as a regressor
# model_with_params <- add_regressor(model_with_params, 'store_nbr')
# model_with_params <- add_regressor(model_with_params, 'family')
# model_with_params <- add_regressor(model_with_params, 'onpromotion')
# model_with_params <- add_regressor(model_with_params, 'city')
# model_with_params <- add_regressor(model_with_params, 'state')
# model_with_params <- add_regressor(model_with_params, 'type.store')
# model_with_params <- add_regressor(model_with_params, 'cluster')
# model_with_params <- add_regressor(model_with_params, 'dcoilwtico')
# model_with_params <- add_regressor(model_with_params, 'locale_name')
# model_with_params <- add_regressor(model_with_params, 'description')
# model_with_params <- add_regressor(model_with_params, 'transferred')
```

```{r, message=FALSE, warning=FALSE}
# # Fit the model
# model_with_params <- fit.prophet(model_with_params, train_data)
# saveRDS(model_with_params, file="model_with_params.RDS")
```

## Load the models 

```{r loading-baseline-model, message=FALSE, warning=FALSE}
model_baseline <- readRDS(file="model_baseline.RDS") 
```

```{r loading-all-f-model, message=FALSE, warning=FALSE}
model_all_features <- readRDS(file="model_all_features.RDS")
```

```{r loading-some-f-model, message=FALSE, warning=FALSE}
model_some_features <- readRDS(file="model_some_features.RDS")
```

```{r loading-params-model, message=FALSE, warning=FALSE}
model_with_params <- readRDS(file="model_with_params.RDS")
```

## Forecast

```{r, message=FALSE, warning=FALSE}
test <- test %>%
  rename(ds = date)
```

## Generate forecast

```{r forecasting, message=FALSE, warning=FALSE}
forecast_baseline <- predict(model_baseline, val_data)
forecast_all_features <- predict(model_all_features, val_data)
forecast_some_features <- predict(model_some_features, val_data)
forecast_with_params <- predict(model_with_params, val_data)
```

## Plot forecast

```{r, message=FALSE, warning=FALSE}
plot_forecast <- function(train_data, forecast_data, title = "Actual vs Forecasted Sales Over Time") {
  # Summarize actual sales from the train dataset
  actual_sales <- train_data %>%
    group_by(date) %>%
    summarize(total_sales = sum(sales, na.rm = TRUE))
  
  # Prepare forecast data for plotting
  forecast_sales <- forecast_data %>%
    select(ds, yhat) %>%
    rename(date = ds, forecasted_sales = yhat) %>%
    group_by(date) %>%
    summarize(total_forecasted_sales = sum(forecasted_sales, na.rm = TRUE))
  
  # Combine actual and forecasted sales into one dataframe
  plot_data <- actual_sales %>%
    full_join(forecast_sales, by = "date")
  
  # Plot actual vs. forecasted sales
  ggplot(plot_data, aes(x = date)) +
    geom_line(aes(y = total_sales), color = "blue", size = 1, na.rm = TRUE) +
    geom_line(aes(y = total_forecasted_sales), color = "red", linetype = "dashed", size = 1, na.rm = TRUE) +
    labs(title = title, x = "Date", y = "Sales") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
}
```

```{r, message=FALSE, warning=FALSE}
plot_forecast(train, forecast_baseline, "Actual vs Baseline Forecasted Sales")
plot_forecast(train, forecast_all_features, "Actual vs All Features Forecasted Sales")
plot_forecast(train, forecast_some_features, "Actual vs Some Features Forecasted Sales")
plot_forecast(train, forecast_with_params, "Actual vs Model with Parameters Forecasted Sales")
```

```{r, message=FALSE, warning=FALSE}
prophet_plot_components(model_baseline, forecast_baseline)
```

```{r, message=FALSE, warning=FALSE}
prophet_plot_components(model_all_features, forecast_all_features)
```

```{r, message=FALSE, warning=FALSE}
prophet_plot_components(model_some_features, forecast_some_features)
```

```{r, message=FALSE, warning=FALSE}
prophet_plot_components(model_with_params, forecast_with_params)
```

## Model Evaluation

```{r, message=FALSE, warning=FALSE}
rmlse <- function(actual, forecast) {
  sqrt(mean((log1p(actual) - log1p(forecast))^2))
}

# Updated evaluation function to include RMLSE
evaluate_forecast <- function(actual, forecast) {
  mae <- mae(actual, forecast)
  mse <- mse(actual, forecast)
  rmse <- rmse(actual, forecast)
  rmlse_value <- rmlse(actual, forecast)
  return(data.frame(MAE = mae, MSE = mse, RMSE = rmse, RMLSE = rmlse_value))
}
```

```{r, message=FALSE, warning=FALSE}
# Extract the actual values from val_data
actual_values <- val_data$y

# Get the forecasted values from each model
forecast_baseline_values <- forecast_baseline$yhat
forecast_all_features_values <- forecast_all_features$yhat
forecast_some_features_values <- forecast_some_features$yhat
forecast_with_params_values <- forecast_with_params$yhat

# Evaluate each model
baseline_metrics <- evaluate_forecast(actual_values, forecast_baseline_values)
some_features_metrics <- evaluate_forecast(actual_values, forecast_some_features_values)
all_features_metrics <- evaluate_forecast(actual_values, forecast_all_features_values)
params_metrics <- evaluate_forecast(actual_values, forecast_with_params_values)
# Compare metrics
comparison <- rbind(baseline_metrics, all_features_metrics, some_features_metrics, params_metrics)
rownames(comparison) <- c("Baseline Model", "All Features Model", "Some Features Model", "Model with Parameters")
print(comparison)
```